Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study

### 2024-01-11 manuscript.Rmd
commit 11/1

Abstract
========
Background
----------
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions. Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods
-------
This was a registry-based study using all patients between 2014 and 2022 from the Karoliniska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference. Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: ...

@@

Introduction
============
Trauma is a leading cause of mortality and morbidity worldwide being the leading cause of death among people under 44 years old in the United States and the fourth leading cause of death in the elderly [1]. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions. Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner rather than only subjective methods where individual cases are reviewed [2]. Examples of opportunities for improvement include lack of resources and management errors.

Albaaj et al. recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna. The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [3]. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.

According to a study on the Korean trauma registry the primary cause of death after trauma was haemorrhage, followed by sepsis/multiorgan failure and central nervous system injury. Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [4]. These are examples of factors that could be important in care on a higher hospital level such as the ICU. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods 
=======
Study design
------------
We conducted single-centre registry-based retrospective cohort study using data from the trauma registry and trauma care quality database at Karolinska University Hospital in Solna. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome of the selected patients was OFI identified by the multidisciplinary review board and then recorded in the trauma care quality database. From 2017 the trauma care quality database includes all patients in the trauma registry as well as the results of the mortality and morbidity review. This project will use these two databases and assess what types of OFI occurred in ICU patients and how different patient factors; such as age, sex, mechanism of injury and injury severity, are associated with OFI using logistic regression. A 5% significance level and 95% confidence levels will be used.

Setting
-------
The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], which means that all patients in Stockholm with severe or suspected severe trauma are sent there. The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6]. All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses. The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for better organization or more senior members taking part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.

Participants
------------
The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and *June 1, 2022 (?)* and who were also admitted to the ICU. *We excluded patients who were younger than 15 years and patients who were dead on arrival.*

Variables and data sources/measurements
---------------------------------------
Study outcome
The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors. *vilka OFIs ska läggas till för att analyseras?*

Patient and process factors 
The factors were chosen from the trauma registry, based on the locally used audit filters ?, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry. The categorical factors were *sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated*. The continuous factors age and Injury Severity Score (ISS) were also included. Working hours was defined as between 8.00 a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.

Bias 
----
*senare*

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------
*senare*

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------
*senare*

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------
Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level. The programming language used was R for all analyses [länk]. The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.

*lägg till?*
A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above. We present sample characteristics using descriptive statistics. Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.

>>>>

Abstract
========
Background
----------
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review. The purpose of these is to identify opportunities for improvement (OFI), which are necessary in order to implement corrective actions. Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods
-------
This was a registry-based study using all patients between 2014 and 2022 from the Karolinska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference. Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: ...

@@

Introduction
============
Trauma is a leading cause of mortality and morbidity worldwide. According to WHO, injury-related deaths constitute 4% of all deaths and take the lives of 4,4 million people yearly worldwide - more than HIV/AIDS, malaria, maternal mortality and tuberculosis combined. The top 3 of 5 causes of death in people aged 5-29 are injury-related [1,2]. Trauma accounts for one-tenth of disability-adjusted life-years (DALYs), most of the burden in low and middle income countries (LMIC) [3, 4]. Trauma costs approximately 3% of GDP yearly globally [5].

A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review. The purpose of these is to identify opportunities for improvement (OFI). Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner [6]. Albaaj et al. recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna. The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [7]. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.

There are many studies on preventable deaths in trauma patients treated in the ICU both regarding to the medical aspect as well as the human aspect. The primary medical causes of death after trauma are haemorrhage, sepsis/multiorgan failure and central nervous system injury. Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [8]. The most prominent OFI in regards to haemorrhage was decision-making compared to errors in technical skill. The OFIs frequently involved the decision between radiology, surgery and further investigation [9].

The University of Pennsylvania ICUs uses telemedicine and telemonitoring to review critical patient events. The areas of improvement they identified were: team dynamics and communication, use of best practices, increased and standardized access to emergency resources and improvement of procedural technique. They recommend other ICUs with telemedicine capacity to adopt a similar strategy in order to present OFIs and educational discussions [10]. Another study looked at how critical information was handled in the ICU and found that laboratory values and test results were the most frequently lost items. This points at the importance of communication and handling of critical information at handoffs of care [11]. From the studies mentioned above some OFIs related to the ICU can be identified: decision-making, access to resources and communication; however, there is still much left to discover. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods 
=======
Study design
------------
?

Setting
-------
The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [12], which means that all patients in Stockholm with severe or suspected severe trauma are sent there. The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [13]. All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses. The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for better organization or more senior members taking part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.

Participants
------------
The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and June 1, 2022 and who were also admitted to the ICU. We excluded patients who were younger than 15 years and patients who were dead on arrival.

Variables and data sources/measurements
---------------------------------------
Study outcome
-------------
The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors...

Patient and process factors 
---------------------------
The factors were chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry. The categorical factors were sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated. The continuous factors age and Injury Severity Score (ISS) were also included. Working hours was defined as between 8.00 a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.

Bias 
----
We used R for all statistical analyses []. A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score... We present sample characteristics using descriptive statistics. The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.

Study size 
----------
<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------
Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI. Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
